A Solver for Generalised Parikh Images of Regular Languages

### 2021-10-10 abstract/main.tex
Some last-minute revisions

\usepackage{textcomp,listings}
\usepackage{biblatex}
\usepackage{mymacros}
\usepackage{mathpartir}
\addbibresource{../bibliography.bib}

@@

\allowdisplaybreaks[1]

\title{A Solver for Generalised Parikh Images of Conjunctions of Regular Languages}
\author{some cool authors}
\institute{Uppsala University, Sweden}

@@

Parikh maps and their image appears naturally as part of many operations in
model checking and the solving of string constraints in automata-based string
solvers such as \Ostrich{} \cite{ostrich}. While it is possible to compute the Parikh image of
an automaton using the method described in \cite{generate-parikh-image}, this
method produces large existentially quantified clauses which are costly to
eliminate and in practice make many real-world problems intractable.
Furthermore, many modern applications require automata with symbolic labels to
handle large alphabets, which has no immediate correspondence in the methods described. Automata-based string solvers like \Ostrich{} also compute
Parikh images on products of automata, derived from conjunctions of string
constraints. Under previous methods this would require the computation of the
product before the Parikh image can be computed, running the risk of an
exponential blow-up.

Addressing these concerns, our approach allows us to extend the computation of
Parikh images to handle symbolic transition labels naturally, while also
allowing us to interleave the computation of arbitrarily deep products of
automata with the Parikh image of their product. This allows us to let both
calculations inform each other, allowing the computation of partial products to
inform the computation of the product and vice versa. Moreover, the scheme
allows us to learn interesting facts (implied clauses) about the problem.

Based on these insights, we present a tool to solve linear constraints on
symbolic automata with counters, amounting to constraints on the Parikh image of
their product. We are able to generate symbolic as well as concrete Parikh images.

@@

A useful example of such a morphism might be computing the length of a string,
which could be recast in terms of the Parikh map by summing the individual
character counts of the vector: $h':\: (\mathbb{Z}^\Sigma, +, \vec{0}) \to
(\mathbb{Z}, +, 0) = \vec{x} \to \sum_{i \in \Sigma} x_i$, where the respective
operators $+$ is element-wise vector addition and integer addition respectively.
This allows us to apply selective logics, symbolic logics, and arithmetic on
counter variabels to the Parikh image calculus.

@@

>>>>

\usepackage{textcomp,listings}
\usepackage{biblatex}
\usepackage{mymacros}
\usepackage{mathpartir}
\usepackage{csquotes}
\usepackage[font={small,it}]{caption}
\addbibresource{../bibliography.bib}

@@

\allowdisplaybreaks[1]
\title{A Solver for Generalised Parikh Images of Regular Languages}
\author{Philipp Rümmer \& Amanda Stjerna}
\institute{Uppsala University, Sweden}

@@

Parikh images appears naturally as part of many operations in model checking or
solving of string constraints in automata-based string solvers such as
\Ostrich{} \cite{ostrich}. While it is possible to compute the Parikh image of
an automaton using the method described in \cite{generate-parikh-image}, this
method produces clauses with many existentially quantified variables which are
costly to eliminate, in practice making many real-world applications
intractable. Furthermore, many applications require automata with symbolic
labels to handle large alphabets like Unicode which would increase the number of
variables even more. Finally, automata-based string solvers compute Parikh
images on products of automata, derived from conjunctions of string constraints.
Using \cite{generate-parikh-image} this would require the up-front computation
of the product before its Parikh image, running the risk of an exponential
blow-up.

Addressing these concerns, our approach allows us to extend the computation of
Parikh images to handle symbolic transition labels ergonomically, while also
allowing us to interleave the computations of the Parikh image of a product and
the product itself. This allows both calculations to inform each other, thereby
eliminating unneccesary work. Moreover, the scheme allows us to learn
interesting facts (in the form of implied clauses) about the problem.

Based on these insights, we present a work-in-progress tool to solve linear
constraints on symbolic automata with counters, amounting to constraints on the
Parikh image of their product. We are able to generate symbolic (\enquote{there are
twice as many \texttt{a}:s as \texttt{b}:s}) as well as concrete Parikh images.

@@

A useful example of such a morphism might be computing the length of a string,
which could be recast in terms of the Parikh map by summing the individual
character counts of the vector: $h':\: (\mathbb{Z}^\Sigma, +, \vec{0}) \to
(\mathbb{Z}, +, 0) = \vec{x} \to \sum_{i \in \Sigma} x_i$. This allows us to
apply e.g. arithmetic on register variables to the Parikh image calculus.

@@

Starting with the automaton shown to the left in figure~\ref{fig:automaton}, we seek
any value in its Parikh image, i.e. a character count.

\begin{figure}[h]
  \caption{Left to right: the starting automaton, the automaton with its corresponding linear equations, and the solution. Note the replacement of the existentially quantified transition variables with character count constants, and the introduction of a linear equation across the edge for \texttt{d} representing a choice between either an~\texttt{a} or a~\texttt{d}. Transitions with a zero value are not used in the solution.}\label{fig:automaton}
  \begin{minipage}[t]{0.3\textwidth}
    \includegraphics[width=\textwidth]{trace-0}
    \end{minipage}
    \begin{minipage}[t]{0.3\textwidth}
    \includegraphics[width=\textwidth]{trace-0-aut-0}
    \end{minipage}
    \begin{minipage}[t]{0.3\textwidth}
      \includegraphics[width=\textwidth]{trace-3-aut-0}
      \end{minipage}
  \end{figure}

  In this case, our calculation will introduce a free integer constant per
  character ($a$, $b$, $c$, and $d$). It will then introduce equations
  preserving flow through the automaton corresponding to how many times that
  transition is used in a run of the automaton. Each transition will be given an existentially
  quantified variable in these equations, and each of our target variables is
  constrained to the sum of the transitions where it occurs as a label. This
  corresponds to the first portion of the formula described
  in~\cite{generate-parikh-image}, but crucially misses the constraints to
  ensure connectedness in the presence of cycles. Instead, we enforce this
  constraint lazily using our own deduction rules.

  Note that this formulation allows us to propagate information between terms of
  a product before materialising it. For many unsatisfiable instances, this
  corresponds to the method recently introduced in~\cite{approximate-parikh},
  where the Parikh image of a product of automata is over-approximated as the
  conjunction of their Parikh images.

Before computation using our rules begin, our prover performs reasoning on the flow equations to simplify them into $d + a = 1 \land c = a$.

We now have a choice of two paths for our run and execute a case split: \textsc{Split}: $a \leq 0$ $\mid$ $a > 0$. The split is performed as close to the initial state as possible and favours deselecting an edge by constraining its corresponding term to be 0.

Since this decision disconnects a loop, the $b$~transition, from the initial state, we evaluate the rule \textsc{Propagate-Connected} to add the constraint $b = 0$. Afterwards we can subsume the propagating machinery for the connectedness constraint, as no more loop transitions can be positive. This leaves us with an automaton with just one nonzero transition, as seen in figure~\ref{fig:automaton}.

As there is just one automaton left, there are no products to compute and this the
calculation is complete. We present the solution $a = b = c = 0, d=1$.
